Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by Galaxy Cluster Abell 2744

### 2015-03-13 snTomas.tex
Finished writing discussion, summary and conclusions.

\title{Illuminating a Dark Lens : A Type Ia Supernova Magnified by Galaxy Cluster Abell 2744}

@@

\begin{abstract}
SN \tomas\ is a Type Ia Supernova (\SNIa) discovered at $z=1.33$
behind the galaxy cluster Abell 2744 ($z=0.308$). This SN has a
projected separation from the cluster core of $\sim$40\arcsec, closer
than any other cluster-lensed \SNIa.  As such, it provides the first
opportunity to confront gravitational lens models of galaxy clusters
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  We derive a tightly constrained measure
of the peak apparent magnitude, corrected for light curve shape and
exinction.  In a cosmology-independent analysis, we find that \tomas\
is $0.71\pm0.12$ magnitudes brighter than unlensed \SNeIa\ at similar
redshift, implying a lensing magnification of $\mu_{\rm
obs}=1.9\pm0.2$.  Predicted magnifications from 7 independent groups
are systematically biased to higher values, with a mean of $\mu_{\rm
mod}=2.4\pm0.3$ and some models disagreeing by $>6\sigma$.  
\textcolor{red}{TODO: summarize lensing discussion.}
%The most
%significant discrepancies arise from models with a strict ``light
%traces mass'' assumption.  We evaluate possible causes for this bias,
%including variation of the mass-to-light ratio in the nearest cluster
%member galaxy.  A sample of \textcolor{red}{$\sim$10??} \SNeIa\ behind
%any single cluster could provide a robust evaluation of systematic
%biases in lens models.  Until such a sample exists, it is advisable to
%increase the estimated uncertainty of any predicted magnification
%by an amount comparable to the dispersion from all available lens
%models.

\end{abstract}

@@

Massive galaxy clusters can be used as cosmic telescopes to magnify
distant background objects through gravitational lensing.  This can
substantially increase the reach of deep imaging surveys, and has
recently been used to discover candidate proto-galaxies formed in the
first Gyr after the Big
Bang \citep{Zheng:2012,Coe:2013} 
\textcolor{red}{(cite other high-z galaxy candidates discovered with lensing)}

Lensed SNe have been discovered
\citep{Goobar:2009,Riehm:2011}

@@

In Section~\ref{sec:DiscoveryAndFollowup} we present the discovery and
follow-up observations of SN \tomas at $z=1.31$, discovered behind the
galaxy cluster Abell 2744.  Sections~\ref{sec:Spectroscopy} and \ref{sec:PhotometricClassification}
describe the spectroscopy and photometry of this SN, leading to a
classification of the object as a normal Type Ia SN.  In
Section~\ref{sec:LensingMagnification} we make a direct measurement of
the magnification of this source due to gravitational lensing, and
compare to predicitions from lens models.  Finally,
Section~\ref{sec:Discussion} discusses the tension between our
magnification measurement and the lens models, with implications for
the systematic error budget in magnification estimates for other
high-$z$ lensed sources.

@@

\begin{figure*}
\begin{center}
%\includegraphics[width=\textwidth]{FIG/discovery_image}
\includegraphics[width=\textwidth]{FIG/discovery_image_lowres}
\caption{  \label{fig:DiscoveryImage} 
SN \tomas\ in the Abell 2744 field.  The left panel shows a
UV/Optical/IR color composite image constructed from all available HST
imaging of the Abell 2744 cluster field.  The inset panels at right
show F814W imaging of the immediate vicinity of SN \tomas,
approximately 40\arcsec\ from the center of the cluster. The top panel
shows the template image, combining all data prior to the SN
appearance.  Labeled ellipses mark the nearest galaxies and their
redshift constraints: a photometric redshift for the most likely host
galaxy is marked in blue, and the spectroscopic redshift for a likely
background galaxy is given in yellow. The bottom panel is constructed
from all HFF F814W imaging taken while the SN was detectable, and
marks the SN location with an arrow.  (Left panel image credit: NASA,
ESA, and J. Lotz, M. Mountain, A. Koekemoer, and the HFF Team (STScI))
}
\end{center}
\end{figure*}

SN \tomas\ was discovered in \HST\ observations with the Advanced
Camera for Surveys (ACS) in the F606W and F814W bands (V and i),
collected on UT 2014 May 15 as part of the Hubble Frontier Fields
(HFF) survey (PI:J.Lotz,
HST-PID:13495).\footnote{\url{http://www.stsci.edu/hst/campaigns/frontier-fields}}
The HFF program is a 3-year director's discretionary initiative that
is collecting 140 orbits of HST imaging (roughly 340 ksec) on six
massive galaxy clusters, plus 6 accompanying ``parallel fields.''
Each field is observed in 3 optical bands (ACS F435W, F606W and F814W)
and 4 infrared (IR) bands (WFC3-IR F105W, F125W, F140W, and F160W),
although the optical and IR imaging campaigns are separated by $\sim$6
months. Abell 2744 was the first cluster observed, with IR imaging
spanning 2013 October-November, and optical imaging from 2014
May-July.  A composite image of the HFF data showing the SN is
presented in Figure~\ref{fig:DiscoveryImage}.  The SN detection was
made in difference images constructed using template imaging of Abell
2744 from HST+ACS observations taken in 2009 (PI:Dupke,
HST-PID:11689).

@@

The most probable host galaxy for SN \tomas\ is a faint and diffuse
galaxy immediately to the south-east of the SN location.  With
photometry of the host galaxy collected from the template images, we
fit the spectral energy distribution (SED) using the {\it BPZ} code --
a Bayesian photometric redshift estimator \citep{Benitez:2000}. The best-fit
SED template match is shown in Figure~\ref{fig:HostSED}.  From the
{\it BPZ} analysis, we found the host to be most likely a late-type
galaxy at a redshift of $z=1.5\pm0.2$.

@@

As in \citet{Patel:2014}, we derive a distance modulus 
from the SALT2 fit using\footnote{This avoids confusion by reserving the symbol $\mu$
to refer to the lensing magnification. The 'dm' is a standard distance
modulus, defined as ${\rm dm}=5\log_{10}d_L+25$, where $d_L$ is the
luminosity distance in Mpc.}

@@

\input{TABLE/lcfits}

The SNANA version of the MLCS2k2 fitter returns a value for the
distance modulus (dm$_{\rm MLCS2k2}$) that has an arbitrary zero point
offset relative to the SALT2 distances (dm$_{\rm SALT2}$).  To put the
two distances onto the same reference frame we add a zeropoint
correction of 0.20 mag to the MLCS2k2 distances as
in \citet{Patel:2014}.  This correction was derived by applying both
fitters to a sample of Type Ia SNe from the SDSS
survey \citep{Holtzman:2008,Kessler:2009b}, with the extinction law
$R_V$ fixed at 1.9.

Final values for the SN distance modulus are shown in
Table~\ref{tab:DistanceAndMagnification}, both for the spectroscopic
redshift ($z=1.31$) and the photometric redshift ($z=1.35$).  The two
light curve fitters are consistent within the uncertainties.

@@

To measure the lensing magnification, we would like to avoid
introducing systematic uncertainties inherent to any assumed
cosmological model \citep[e.g.][]{Nordin:2014}.  To that end, we
follow \citet{Patel:2014} and define the magnification by comparing
the measured distance modulus of SN \tomas\ against an average
distance modulus derived from a ``control sample'' of unlensed Type Ia
SN at similar redshift.  This allows us to make only the minimal
assumption that the redshift-distance relationship for Type Ia SN is
smooth and approximately linear over a small redshift span, which
should be true for any plausible cosmological model.

The unlensed sample comprises 18 spectroscopically confirmed Type Ia
SNe in the range $1.14\leq z \leq1.42$ from the GOODS and SCP
surveys,\footnote{GOODS: Great Observatories Origins Deep Survey,
PI:Giavalisco, HST-PID:; SCP: Supernova Cosmology Project,
PI:Perlmutter} which used the HST Advanced Camera for
Surveys \citep{Riess:2006,Suzuki:2012}.  Using the SALT2 and MLCS2k2
fitters as described above, we get distance modulus measures for every
object in this control sample.  We then fit a linear relationship for
distance modulus vs. redshift, and derive a prediction for the
distance modulus of a normal Type Ia SN at the redshift of SN \tomas
(Figure~\ref{fig:MagnificationMeasurement}).
This predicted value is given in
Table~\ref{tab:DistanceAndMagnification} under the ``Control'' column.
The difference between the observed distance of SN \tomas\ and this
control sample value is attributed to the magnification from
gravitational lensing:

@@

\noindent This inferred magnification is reported in the final column of
Table~\ref{tab:DistanceAndMagnification}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/magnification_measurement}
\caption{ Measurement of the lensing magnification from comparison of
the \tomas\ distance modulus to a sample of unlensed field SN.
The distance modulus for each SN is derived from light curve fits
using the SALT2 fitter (top panel) and the MLCS2k2 fitter (right).
\label{fig:MagnificationMeasurement} }
\end{center}
\end{figure}

@@

1) All of the lens models predict a higher magnification than we observe, some with discrepancies >5 sigma.  This suggests a systematic bias inherent to all the models -- at least for this particular line of sight.

2) the addition of new lensing constraints from the HFF data does not reduce the tension.  From both of the most recent models (Jauzac+ 2014 and Lam+ 2014) the predicted magnification is still significantly higher than the observed.

3) The pre-HFF models that come closest to matching the observed magnification are the Zitrin-NFW and Merten models (also the Williams model, though it has very large uncertainties).  Those happen to be two models that relax or remove the assumption of light-traces-mass.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/lensing_test}
\caption{ 
Comparison of the observed lensing magnification to predictions from
lens models.
\label{fig:lensingtest} }
\end{center}
\end{figure}

\section{Discussion}

Possible sources for the bias.

Implications for the systematic error budget of high-z lenses.

Design elements for a survey to build a bigger sample : HST GO, Snapshot, ground-based.

>>>>

\title{Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by Galaxy Cluster Abell 2744}

@@

\begin{abstract}
SN \tomas\ is a Type Ia Supernova (\SNIa) discovered at $z=1.31$
behind the galaxy cluster Abell 2744 ($z=0.308$). This SN has a
projected separation from the cluster core of $\sim$40\arcsec, closer
than any other cluster-lensed \SNIa.  As such, it provides the first
opportunity to confront gravitational lens models of galaxy clusters
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  We derive a tightly constrained measure
of the peak apparent magnitude, corrected for light curve shape and
exinction.  In a cosmology-independent analysis, we find that \tomas\
is $0.75\pm0.10$ magnitudes brighter than unlensed \SNeIa\ at similar
redshift, implying a lensing magnification of $\mu_{\rm
obs}=2.00\pm0.19$.  We compare this direct measurement of the
magnification against predicted magnifications from 11 lens models,
representing a broad range of methodologies and incorporating a
variety of strong- and weak-lensing constraints.  The models are
collectively fairly accurate, with a naive unweighted mean of
$\mu_{\rm mod}=2.6$.  However, they are systematically biased to
higher values: none of the tested models predict a magnification equal
to or lower than the observed $\mu$, and some models are more than
$5\sigma$ above the measured magnification.  We evaluate possible
causes for this bias, and find that intercomparison of the lens models
rules out several simple possibilities.  A larger sample of
$\sim$10 \SNeIa\ behind any single cluster could provide a robust
evaluation of systematic biases in lens models.  Until such a sample
exists, it is advisable to increase the estimated uncertainty of any
predicted magnification by an amount comparable to the dispersion from
all available lens models.
\end{abstract}

@@

\textcolor{red}{TODO: write and polish introduction}.
Massive galaxy clusters can be used as cosmic telescopes to magnify
distant background objects through gravitational lensing.  This can
substantially increase the reach of deep imaging surveys, and has
recently been used to discover candidate proto-galaxies formed in the
first Gyr after the Big 
Bang \citep{Zheng:2012,Coe:2013} 
\textcolor{red}{(cite other high-z galaxy candidates discovered with lensing)}

\textcolor{red}{TODO: describe past examples of lensed SN studies  \citep{Goobar:2009,Riehm:2011,Quimby:2014}.}

@@

In Section~\ref{sec:DiscoveryAndFollowup} we present the discovery and
follow-up observations of SN \tomas at $z=1.31$, discovered behind the
galaxy cluster Abell 2744.  Sections~\ref{sec:Spectroscopy} and \ref{sec:PhotometricClassification}
describe the spectroscopy and photometry of this SN, leading to a
classification of the object as a normal Type Ia SN.  In
Section~\ref{sec:Magnification} we make a direct measurement of
the magnification of this source due to gravitational lensing, and
compare to predictions from lens models.  Finally,
Section~\ref{sec:Discussion} discusses the tension between our
magnification measurement and the lens models, with implications for
the systematic error budget in magnification estimates for other
high-$z$ lensed sources.

@@

\begin{figure*}
\begin{center}
%\includegraphics[width=\textwidth]{FIG/discovery_image}
\includegraphics[width=\textwidth]{FIG/discovery_image_lowres}
\caption{  \label{fig:DiscoveryImage} 
SN \tomas\ in the Abell 2744 field.  The left panel shows a
UV/Optical/IR color composite image constructed from all available HST
imaging of the Abell 2744 cluster field.  The inset panels at right
show F814W imaging of the immediate vicinity of SN \tomas,
approximately 40\arcsec\ from the center of the cluster. The top panel
shows the template image, combining all data prior to the SN
appearance.  Labeled ellipses mark the nearest galaxies and their
redshift constraints: a photometric redshift for the most likely host
galaxy is marked in blue, and the spectroscopic redshift for a
background galaxy is given in yellow. The bottom panel is constructed
from all HFF F814W imaging taken while the SN was detectable, and
marks the SN location with an arrow.  (Left panel image credit: NASA,
ESA, and J. Lotz, M. Mountain, A. Koekemoer, and the HFF Team (STScI))
}
\end{center}
\end{figure*}

SN \tomas\ was discovered in Hubble Space Telescop (\HST) observations
with the Advanced Camera for Surveys (ACS) in the F606W and F814W
bands (V and i), collected on UT 2014 May 15 as part of the Hubble
Frontier Fields (HFF) survey (PI:J.Lotz,
HST-PID:13495).\footnote{\url{http://www.stsci.edu/hst/campaigns/frontier-fields}}
The HFF program is a 3-year director's discretionary initiative that
is collecting 140 orbits of HST imaging (roughly 340 ksec) on six
massive galaxy clusters, plus 6 accompanying ``parallel fields.''
Each field is observed in 3 optical bands (ACS F435W, F606W and F814W)
and 4 infrared (IR) bands (WFC3-IR F105W, F125W, F140W, and F160W),
although the optical and IR imaging campaigns are separated by $\sim$6
months. Abell 2744 was the first cluster observed, with IR imaging
spanning 2013 October-November, and optical imaging from 2014
May-July.  A composite image of the HFF data showing the SN is
presented in Figure~\ref{fig:DiscoveryImage}.  The SN detection was
made in difference images constructed using template imaging of Abell
2744 from HST+ACS observations taken in 2009 (PI:Dupke,
HST-PID:11689).

@@

The most probable host galaxy for SN \tomas\ is a faint and diffuse
galaxy immediately to the south-east of the SN location.  With
photometry of the host galaxy collected from the template images, we
fit the spectral energy distribution (SED) using the {\it BPZ} code --
a Bayesian photometric redshift estimator \citep{Benitez:2000}. The
best-fit SED template match is shown in Figure~\ref{fig:HostSED}.
From the {\it BPZ} analysis, we found the host to be most likely an
actively star-forming galaxy at a redshift of $z=1.5\pm0.2$.  Another
nearby bright galaxy to the East of the SN has a spectroscopic
redshift of $z=1.742$, making it most likely a background object.
This redshift was determined from a spectrum taken with the G141 grism
of the HST WFC3-IR camera, collected as part of the Grism Lens
Amplified Survey from Space (GLASS, PI:Treu, PID:13459).

@@

As in \P14, we derive a distance modulus\footnote{We use 'dm' to
indicate the distance modulus to avoid confusion, reserving the symbol
$\mu$ to refer to the lensing magnification. This 'dm' is a standard
distance modulus, defined as ${\rm dm}=5\log_{10}d_L+25$, where $d_L$
is the luminosity distance in Mpc.}  from the SALT2 fit using

@@

\input{TABLE/dist_mag}

The SNANA version of the MLCS2k2 fitter returns a value for the
distance modulus (dm$_{\rm MLCS2k2}$) that has an arbitrary zero point
offset relative to the SALT2 distances (dm$_{\rm SALT2}$).  To put the
two distances onto the same reference frame we add a zeropoint
correction of 0.20 mag to the MLCS2k2 distances as
in \P14.  This correction was derived by applying both
fitters to a sample of Type Ia SNe from the SDSS
survey \citep{Holtzman:2008,Kessler:2009b}, with the extinction law
$R_V$ fixed at 1.9.

Final values for the SN distance modulus are shown in
Table~\ref{tab:MeasuredMagnification}, both for the spectroscopic
redshift ($z=1.31$) and the photometric redshift ($z=1.35$).  The two
light curve fitters are consistent within the uncertainties.

@@

To measure the lensing magnification, we would like to avoid
introducing systematic uncertainties inherent to any assumed
cosmological model \citep[e.g.][]{Nordin:2014}.  To that end, we
follow \P14\ and define the magnification by comparing
the measured distance modulus of SN \tomas\ against an average
distance modulus derived from a ``control sample'' of unlensed Type Ia
SN at similar redshift.  This allows us to make only the minimal
assumption that the redshift-distance relationship for Type Ia SN is
smooth and approximately linear over a small redshift span, which
should be true for any plausible cosmological model.

The unlensed sample comprises 18 spectroscopically confirmed Type Ia
SNe in the range $1.14\leq z \leq1.42$ from the GOODS and SCP
surveys,\footnote{GOODS: Great Observatories Origins Deep Survey,
PI:Giavalisco, HST-PID:; SCP: Supernova Cosmology Project,
PI:Perlmutter} which used the HST Advanced Camera for
Surveys \citep{Riess:2006,Suzuki:2012}.  Using the SALT2 and MLCS2k2
fitters as described above, we get distance modulus measures for every
object in this control sample.  We then fit a linear relationship for
distance modulus vs. redshift, and derive a prediction for the
distance modulus of a normal Type Ia SN at the redshift of SN \tomas
(Figure~\ref{fig:MagnificationMeasurement}).
This predicted value is given in
Table~\ref{tab:MeasuredMagnification} under the ``Control'' column.
The difference between the observed distance of SN \tomas\ and this
control sample value is attributed to the magnification from
gravitational lensing:

@@

\noindent This inferred magnification is reported in the final column of
Table~\ref{tab:MeasuredMagnification}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/snTomas_hubble_diagram}
\caption{ Measurement of the lensing magnification from comparison of
the \tomas\ distance modulus to a sample of unlensed field SN.
The distance modulus for each SN is derived from light curve fits
using the MLCS2k2 fitter (top panel) and the SALT2 fitter (bottom panel).
\label{fig:MagnificationMeasurement} }
\end{center}
\end{figure}

\section{Discussion}
\label{sec:Discussion}

@@

Before the Frontier Fields observations began, the Space Telescope
Science Institute (STScI) issued a call for lens modeling teams to
generate mass models of all 6 Frontier Field clusters, using a shared
collection of all imaging and spectroscopic data available at the
time.  In response to this opportunity, five teams generated eight
models for Abell 2744.  These models were necessarily quite
preliminary, and they were principally intended to be used for quick
estimation of magnifications for lensed background sources. An
interactive web tool was created by D. Coe and hosted at STScI, to
extract magnification estimates and uncertainties from each model for
any given redshift and position.  In this work we also consider three
additional models that were created later, taking advantage of new
multiply-imaged galaxies discovered in the HFF imaging as well as new
redshifts for lensed background galaxies.

\input{TABLE/magnifications}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/snTomas_magnifications}
\caption{ 
Comparison of the observed lensing magnification to predictions from
lens models. Solid vertical lines show the constraints from SN \tomas\
derived in Section~\ref{sec:Magnification} using the SALT2 (red) and
MLCS2k2 (blue) fitters, with shaded regions marking the total
uncertainty for each. Markers with horizontal error bars show the
median magnification and 68\%\ confidence region from each of the 10
lensing models.  Circles indicate models that use only strong-lensing
constraints, while diamonds denote those that also incorporate
weak-lensing measurements.  Models shown with open markers were
constructed using only data available before the start of the Frontier
Fields observations.  Filled markers indicate models that use
additional input constraints, including new multiply imaged systems
and redshifts.  The seven models shown in black at the top are from
the ``parametric'' family, and the four in green at the bottom are
``free-form.''  The dashed grey line marks the unweighted mean of all
model magnifications, with grey shading indicating the standard error
of the mean from naively assuming all are independent predictions.
\label{fig:lensingtest} }
\end{center}
\end{figure}

Table~\ref{tab:PredictedMagnifications} lists the 11 models, giving
for each one the predicted magnification and uncertainty for a source
at $z=1.31\pm0.01$ and at the position of \tomas.  These
models represent a broad sampling of the techniques and assumptions
that can be applied to the modeling of mass distributions in galaxy
clusters. A rigorous comparison of these disparate lens modeling
techniques is beyond the scope of this work. Table 10 includes a brief
description of each model, but for a complete discussion of the
methodology the reader is referred to the listed references.

In Figure~\ref{fig:lensingtest} the model predictions are plotted
alongside the observed magnification of SN \tomas, derived in
Section~\ref{sec:Magnification}.  This comparison shows that these 11
models are largely consistent with each other. Naively treating each
model as an independent prediction for the magnification (and ignoring
their quoted uncertainties), one would get a mean for the full sample
of $\mu=2.62\pm0.34$ (shown as a grey band in
Figure~\ref{fig:lensingtest}).  This is separated from the observed SN
magnification by $\delta\mu/\mu=30\%$, which is a 1.7$\sigma$
difference.

The models shown in Figure~\ref{fig:lensingtest} are separated along
the $y$ direction into two broad classes, with the six {\it
parametric} models on top and the four {\it free-form} models on the
bottom.  Broadly speaking, parametric models use parameterized density
distributions to describe the arrangement of mass within the cluster,
typically associating massive dark matter halos with the positions of
bright cluster member galaxies. Therefore, parametric models rely (to
varying degrees) on the assumption that the observed light from
cluster galaxies is a good tracer of the cluster's dark matter.
Free-form models divide the cluster into a grid, generally using an
adaptive mesh to get better sampling in denser regions. Each grid cell
is assigned a mass, and then the grid spacing and the mass values are
iteratively refined to match the observed lensing constraints.

Figure~\ref{fig:lensingtest} also separates the models based on the
scope of input data constraints used.  Seven of the models rely only
on strong-lensing constraints (plotted as circles), while the other
four also use weak-lensing measurements (diamonds).  Eight of the
models were constructed using only pre-HFF data (open symbols), and
three others added new multiply-imaged systems and redshifts.

In spite of the great variety in input data and methodology, these
lens models overall are delivering consistent and fairly accurate
estimates of the magnification.  Furthermore, the scatter amongst
models provides a reasonable estimate of the magnification
uncertainty.  This is largely in agreement with the results
of \citetalias{Patel:2014} and \citet{Nordin:2014}.  The fact that the
model predictions are collectively within $2\sigma$ of the observed
magnification is especially encouraging for these Abell 2744
models. This is a merging cluster with a complex mass distribution,
and most of the models we are evaluating are preliminary products,
generated quickly and without access to the rich data from the
Frontier Fields program.

However, beyond this first-order agreement, there is a systematic bias
apparent. All of the lens models predict a magnification that is {\it
higher} than the observed value, and no model has a median value that
falls within the $1\sigma$ range of the observed magnifications.  It
is important to emphasize that SN \tomas\ only samples a single
sight-line through the cluster, so any conclusions to be drawn from
this analysis are necesssarily limited.  Nevertheless, a systematic shift
common to all models is surprising, given the wide range of modeling
strategies, input data, and physical assumptions represented by this
set of models.  In the following subsections we examine possible
explanations for this apparently universal bias.

\subsection{Nearby Cluster Member Galaxy}
\label{sec:NearbyClusterMemberGalaxy}

The line of sight to SN \tomas\ is 5\farcs8 from a bright cluster
member galaxy, due North of the SN position (see
Figure~\ref{fig:DiscoveryImage}).  If the mass-to-light ratio (M/L)
for this galaxy were significantly different from the M/L for other
cluster member galaxies, then its proximity to the SN sight-line could
introduce a bias in the magnification.  This bias might be
particularly acute for models that make a strict ``light traces mass''
assumption, such as the Sharon, CATS, Jauzac, GLAFIC and Zitrin-LTM
models.

We tested this hypothesis using the Jauzac model by allowing the mass
of the nearby cluster member galaxy to vary as a free parameter in the
model. We found that the change in the SN \tomas\ magnification
prediction was less than $\Delta\mu=0.1$.  This additional dispersion
is already included in the uncertainties quoted for that model in
Table~\ref{tab:PredictedMagnifications} and
Figure~\ref{fig:lensingtest}.  Furthermore, an erroneous M/L
value for the nearby cluster member galaxy could not explain the
systematic shift of all lens models, because some do not incorporate
cluster member galaxies into their constraints at all (the free-form
Bradac, Williams and Merten models).

\subsection{Sparse Strong Lensing Constraints}
\label{sec:SparseStrongLensingConstraints}

At the outset of the HFF survey, only 33 multiply-imaged galaxies
behind Abell 2744 were known.  One might expect that the addition of
new lensing constraints from the HFF data would improve the model
constraints and reduce the tension between the predicted and observed
magnification.  However, for both of the most recent
models \citep{Jauzac:2014c,Lam:2014} that include new strong-lensing
constraints, the predicted magnification is still significantly higher
than the observed value.  Again the CATS model family provides a
useful case study.  The Jauzac model shares the same lens modeling
software (LensTool), and the same primary methodology as the earlier
CATS models.  A key difference is that the Jauzac model incorporates
many new multiply imaged galaxies and corrects a mis-identified
multiply-imaged system near the SN \tomas\ position.  These
improvements should in principle make the Jauzac model magnification
estimates more accurate, though we in fact see the opposite result:
the Jauzac model predicts a $\mu$ that is higher than the measured
value by $5.6\sigma$.

This shift to higher magnifications is not, however, unique to the
SN \tomas\ position.  \citet{Jauzac:2014c} noted that their updated
mass model of Abell 2744 results in a systematic increase in the
magnification values across the cluster field. Examining a sample of
$\sim$30 multiple images, \citeauthor{Jauzac} found that the
magnifications increased by a factor of $\sim1.5$ relative to pre-HFF
models.  Although this single sight-line can only provide a very
limited test of the model, the $5.6\sigma$ discrepancy is at least a
strong suggestion that improving the number and quality of
strong-lensing constraints can still lead to magnification maps that
are susceptible to systematic biases.

\subsection{Mass Profile Truncation}
\label{sec:MassProfileTruncation}

Seven of the mass models evaluated here are constrained only using
strong-lensing features such as multiply-imaged background galaxies
and highly magnified arcs.  There are now $\sim$150 known lensed
images behind Abell 2744 \citep{Jauzac:2014c}, but SN \tomas\ is
located several arcseconds outside the core region of the cluster
where these multiply-imaged galaxies are found.  Therefore these seven
``strong-lensing only'' models must necessarily rely on extrapolations
to provide a prediction for the SN \tomas\ magnification.  If the true
cluster density profile for Abell 2744 happens to be truncated -- with
a sharp drop right at the edge of this core strong-lensing region --
then these seven models might well overestimate the mass interior to
the \tomas\ position, and thus systematically overestimate the
magnification.

The other four models in our comparison set also use the strong
lensing constraints from the core region, but also utilize weak
lensing measurements to provide additional constraints on the mass
distribution farther from the cluster core.  The signal from weak
lensing relies on a large sample of background galaxies, so this
constraint operates principally at separations more than 1\arcmin from
the cluster core.  At a projected separation of $\sim40\arcsec$,
SN \tomas\ falls in between the strong- and weak-lensing regimes, and
one would expect that models incorporating both of those constraints
would be less susceptible to the systematic bias of a truncated mass
profile.  The lowest magnification estimate in our sample (from the
Merten model) is among this sub-sample.  However, we find that
collectively these four strong+weak models (shown as diamonds in
Figure~\ref{fig:lensingtest}) exhibit the same propensity to
overestimate the magnification along this line of sight.  Thus, a
sudden change in the mass profile outside the strong lensing region is
not a likely explanation for this small systematic bias.

With two versions of the CATS model, we also have a more direct test
of the effect of introducing weak lensing constraints.  The initial
CATS model uses only strong-lensing constraints, and gives
$\mu=2.27^{+0.05}_{-0.04}$, slightly higher than the measured value.
A second iteration of this model, labeled here as CATS-SL+WL, used the
same strong-lensing features, but added in weak lensing constraints.
The revised magnification of $\mu=2.62\pm0.18$ is further from the
measured value, which serves to reinforce the conclusion that adding
weak lensing constraints cannot resolve this bias.

\section{Summary and Conclusions}
\label{sec:SummaryAndConclusions}

The appearance of a Type Ia SN behind a massive galaxy cluster
provides a rare opportunity to use a standard candle for a direct
measurement of the absolute magnification due to gravitational
lensing.  The discovery of SN \tomas\ in the HFF imaging of Abell 2744
offers the first chance to apply this test on a cluster with multiple
publicly-available lens models.  We have found that the spectrum and
light curve of SN \tomas\ are well matched by templates of a normal
Type Ia SN at $z=1.31\pm0.01$.  Using the two most prevalent \SNIa\
light curve fitters, SALT2 and MLCS2k2, we get a consistent
measurement of the distance modulus
(Table~\ref{tab:MeasuredMagnification}).  Using a
cosmology-independent comparison against a sample of unlensed \SNeIa\
at similar redshifts, we find that SN \tomas\ is $\sim0.7$ magnitudes
brighter than the field sample would predict.  Attributing this
difference to the gravitational lensing magnification (and accounting
for the intrinsic scatter in luminosity of the \SNIa\ population), we
have derived a consistent measured magnification of $\mu_{\rm
SALT2}=1.89\pm0.29$.  $\mu_{\rm MLCS2k2}=2.00\pm0.19$, from the two
light curve fitters.

Taking advantage of the availability of eleven well-constrained lens
models for the Abell 2744 cluster, we have used SN \tomas\ to ask
how accurately these lens models can predict the magnification along
this line of sight.   We find that these models are consistent, and
fairly accurate, collectively predicting $\mu=2.60\pm0.34$.  This is
encouraging, and reinforces the quality and value of these public lens
models for studying magnified background objects.   However, the fact
that all models predict a larger magnification value than we observe
is an indication that there is a small systematic bias inherent to
this cluster or sight-line.

We have speculated on the possible origin of this systematic bias,
evaluating three possible causes:

\begin{enumerate}
\item The close proximity of a cluster member galaxy is leading to a
biased magnification, because that galaxy's M/L is atypical.
\item The models are limited by a scarcity of strong lensing constraints
\item The cluster mass profile exhibits a sudden change outside of the core
strong-lensing region, where we lack multiply-imaged background
sources to constrain it. 
\end{enumerate}

\noindent None of these three explanations is completely
satisfactory.  For each of these scenarios, we would expect that
certain subsets of the available lens models would be less sensitive
to the proposed systematic bias.  However, the small discrepancy
between the predicted and observed lensing magnification is persistent
when we bifurcate the sample of lens models according to methodology
(parametric vs. free-form), number of strong-lensing constraints (pre-
vs post-HFF), and scope of lensing constraints (strong-lensing only
vs. strong+weak).  We do not have any definitive explanation for this
magnification discrepancy.

This single object behind a single cluster is not in and of itself a
cause for alarm.  Previous analyses of lensed \SNIa\ found no
significant discrepancy beween the observed \SNIa\ magnifications and
the predictions from lens models
(\citetalias{Patel:2014}; \citealt{Nordin:2014}), albeit with a much
smaller set of lens models being tested.  The observed systematic bias
for \tomas\ is small, and many of the lens models being evaluated are
preliminary models that have not been updated to include all of the
HFF data.  Future revisions of the lens models for Abell 2744 should
either incorporate the observed magnification of \tomas\ as a new
model constraint, or can revisit this test to evaluate whether the
bias persists.

Increasing the sample of SNe behind well studied strong lensing
clusters would allow this test to be repeated and refined.  With 10 or
100 such objects, it would be possible to see whether the SN \tomas\
$\mu$ discrepancy is simply an outlier, or an indication of a more
pernicious systematic error.  Any cluster that has been vetted by
pencil-beam magnification tests using lensed \SNIa\ will be able to
provide a more reliable measure of the magnifications for very high
redshift objects.  Similarly, a broad sample of SNe like \tomas\ would
help to define the preferred lens modeling methodology by highlighting
any models that consistently perform well in \SNIa\ lensing tests.
The ongoing FrontierSN program will discover and follow any more
highly magnified SNe that appear behind the Frontier Field
clusters. Unfortunately, the HFF survey is not designed with high-z
transient discovery as a primary science goal, so the FrontierSN
effort will likely add no more than 1-3 new lensed \SNIa. Further
imaging of strong-lensing clusters with \HST\ or the James Webb Space
Telescope (JWST) could enable a larger sample to be collected,
especially if the filters and cadence are optimized for detection
of \SNIa\ at $z>1$.  Massive clusters such as Abell 2744 will continue
to be attractive as cosmic telescopes, allowing the next generation of
telescopes to reach the faintest objects in the very early universe.
The puzzling bias revealed by SN \tomas\ supports a concerted effort
to improve these lenses with a larger sample of magnified SNe.
